---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# droughtevents

<!-- badges: start -->
<!-- [![R-CMD-check](https://github.com/ajpelu/droughtevents/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/ajpelu/droughtevents/actions/workflows/R-CMD-check.yaml) -->
<!-- [![Codecov test coverage](https://codecov.io/gh/ajpelu/droughtevents/branch/main/graph/badge.svg)](https://app.codecov.io/gh/ajpelu/droughtevents?branch=main) -->
<!-- badges: end -->

## Overview

`droughtevents` provides tools to detect, summarize, and visualize drought
events from drought-related index time series (e.g., SPEI, SPI). Given a
time series and a threshold, the package identifies periods of sustained
below-threshold values, computes summary statistics for each event
(duration, intensity, severity, timing), and offers `ggplot2`-based
plotting functions to visualize the time series and highlight the
detected events.

## Installation

You can install the development version of `droughtevents` from
[GitHub](https://github.com/) with:

``` r
# install.packages("remotes")
remotes::install_github("ajpelu/droughtevents")
```

## Usage

```{r example}
library(droughtevents)
library(ggplot2)
```

### Detect drought events

`droughts()` identifies drought events in a time series when a given
index falls below a specified threshold for at least `min_duration`
consecutive months (2 by default).

```{r detect-droughts}
data(spei_granada)

result <- droughts(spei_granada, vname = "spei12", threshold = -1.28)

result$drought_assessment
```

The returned object is a named list with three elements:

- `data`: the original data, with drought flags and durations added.
- `drought_events`: only the rows that belong to a detected drought event.
- `drought_assessment`: one row per event, with its duration, intensity,
  severity, and timing.

### Plot the time series

`plot_drought_ts()` draws the index as a bar plot, coloring positive
(wet, *blue*) and negative (dry, *red*) periods differently.

```{r plot-ts}
plot_drought_ts(spei_granada, vname = "spei12", title = "SPEI-12 Time Series")
```

You can also aggregate by year:

```{r plot-ts-by-year}
plot_drought_ts(spei_granada, vname = "spei12", by_year = TRUE)
```

### Highlight drought events on the plot

`add_drought_events()` overlays the detected events on top of a plot
created with `plot_drought_ts()`, as shaded bands or vertical lines, or
both. 

```{r plot-with-events}
p <- plot_drought_ts(spei_granada, vname = "spei12", title = "SPEI-12 Time Series")

p |>
  add_drought_events(
    drought_assessment = result$drought_assessment,
    which_events = "top",
    metric = "severity",
    top_n = 5,
    type = "both",
    show_severity = FALSE
  )
```

## Data

The package ships with `spei_granada`, a monthly SPEI time series
(6, 12, 24, and 48-month scales) for Granada, Spain, covering 1901-2024.
See `?spei_granada` for details and the data source.

## References

Vicente-Serrano, S.M., Beguería, S., López-Moreno, J.I. (2010). A
Multiscalar Drought Index Sensitive to Global Warming: The Standardized
Precipitation Evapotranspiration Index. *Journal of Climate*, 23(7),
1696-1718. <https://doi.org/10.1175/2009JCLI2909.1>

## License

<!-- Add your license here, e.g. MIT, GPL-3 -->
